Characterizing subarctic biomes for land surface modeling of pollution and climate risk

### 2021-03-22 VegOzone.tex
Second version of conclusions - not spell ckecked

\begin{abstract}
Here we assess the importance of characterization of key meteorological variables and air pollution profiles on the modeled susceptibility of subarctic vegetation to ozone. Changes in climate (e.g. temperature, water availability) will manifest in both direct and indirect effects influencing the species' sensitivity to air pollution. A likely increase in frequency of persistent heatwaves in the future, will inflict heat related stress while generating favorable conditions for the formation of ambient air ozone, e.g. due to wildfires.
We have monitored ambient air ozone concentrations at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in Northern Norway in the vegetation period 2018/19. Due to data acquisition problems, ozone concentrations for two weeks in July 2018 were missing from our record. We present a reconstruction based on probability density function with respect to the Swedish and Finnish atmospheric monitoring sites and compare with available surface ozone products for the region. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged substantially from the multi annual mean precipitation varied only to some extend. 
We investigate the difference in uptake of ozone using the $\mathrm{DO_3SE}$ model, for typical vegetation at the location. We find that bespoke parameterization for subarctic vegetation is essential for an accurate estimation of Phototoxic Ozone Dose (POD) and an assessment of damage risk. Accurate estimates of the start and length of growing season are also crucial.
We find that temperature and its derivative vapor pressure deficit are likely the most important climate variables while the variability of ozone concentration within climatologically reasonable limits plays a minor role.
The use of subarctic parameterization suggests an improvement in risk assessment as POD values are increased in 2018; corroborated by evidence of visible injury in the Svanhovd ozone garden.
\end{abstract}

@@

To measure the risk of damage inflicted to vegetation through ozone, different metrics have been defined. Two main types can be differentiated, concentration-based and exposure-based metrics, e.g., Accumulated Over a Threshold $\mathrm{x}$ ($\mathrm{AOT_x}$) and Phyto-toxic Ozone Dose over a threshold $\mathrm{y}$ ($\mathrm{POD_y}$). Due to their simplicity, concentration-based metrics are still widely used, despite comparisons between the two methods (e.g. $\mathrm{AOT_{40}}$ with $\mathrm{POD_3}$) show that exposure-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the more climatically extreme parts of Europe \citep{GCB:Mills2010,ICP:MappingManual2017}. Most of these previous studies focused on the Mediterranean region.

@@

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated onset of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm}.

@@

\conclusions[Discussion and conclusions]
\label{sec:conc}
We have examplarily studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. We also assessed implications on potential ozone damage risks arising at the advant of climate change. We found that conditions for ozone formation were more favourable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occured more frequently and at higher concentration levels in 2018. This was in particular caused by the extended heatwave in spring and early summer over whole Fennoscandia and associated, extensive wildfires esecially in Central Sweden (mid--end July).

The timing and trend in the onset of growing season may differ substantially between individual species in comparison to the thermal or latitudinal estimated growing season. Species-specific methods to estimate the onset \citep{IVL:Karlsson2018} show a later onset for European silver birch and earlier for Norway spruce, compared to the thermal growing season estimate. Based on MODIS data, we estimated the growing season in 2018 to be at least one month shorter than in 2019 for Norway spruce, and half a month shorter for downy birch and perennial grassland, respectively.

We have proposed a routine for gap filling of ozone data following a Reynolds's decomposition into annomalies and climatology (mulit-annual mean). We also assessed the quality of available global (ECWMF MACC and CAM reanalysis) and regional (Copernicus) near-surface ozone concentrations products and conclude that only the Copernicus product can sufficiently reproduce observed ozone concentrations in northern Fennoscandia.

Although high and more frequent ozone peak concentrations were observed in 2018, ozone concentrations were not significantly elevated by statistical means. It therefor occurs that weather conditions and constantly high ozone concentration are more important in risk assessment using flux-based metrices like $\mathrm{POD_y}$ than episodes of peak concentrations over shorter periodes of time. Consequently, accute cellular damage caused by a very high fluxes over a short periode of time is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. higher susceptibility to ozone depending on leaf age, in the associated damage functions might give a more differenciated picture.

Besides a currently missing coupling to photosynthesis or nutrient cycle, ozone inflicted damage on stomatal conductance is also not accounted for in the $\mathrm{DO_3SE}$ model. Cumulative ozone uptake has been reported to reduce stomatal conductance significantly \citep[e.g.]{BGS:Lombardozzi2012} and subsequently reduces uptake of ozone. On the other hand, an ozone induced sluggishness \citep{SR:Hoshika2015} renders stomata inapt to fully close and would lead to increased ozone uptake. The presented $\mathrm{POD_y}$ may therefor be regarded as an upper limit.
{\bf TODO: might be interesting to look at Mills et al 2003 and the different spatial extents of risk estimated when using AOT40 vs PODy. That could be useful to explain why we focus on PODy (rather then AOT40).- Lisa Come back to this later on in the discussions(?) - it has not been breached at Svanvik in northern Fennoscandia.}

We defined hypothetical temperature response functions representing different levels of species assumed acclimation to cold climates. We assume the acclimation to increase the enclosed area between probability density function and the respective response function. These hypothetical species or PFTs still need varification by experiemental data. Our results, however, allowed for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In particular, we identified the following main effects on the model predicted ozone uptake with respect to the default parameterization used in risk assessment:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Predicted ozone uptake with bespoke temperature and light response function and length and timing of growing season is larger than for default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species, especially for downy birch and perennial grassland which have been studied less intensively than Norway spruce which showes a lower susceptibility to ozone \citep{Amb:Girgzdiene2009}.

Open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch were exposed elevated ozone concentrations showed biomass reductions of both downy birch and mountain birch seedlings after one-season exposure \citep{Amb:Manninen2009}. The leaves of the mountain birch in the high ozone treatments showed necrotic stipples as early as June. As mountain birch very rapidly develops during spring and early summer, compensation for damaged leaves is difficult for this species. In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level \chem{O_3} impacts after three years of exposure. There is no field data comparing the ozone sensitivity of birches and Scots pine from northern Europe.

@@

Beyond the risk assessment of ozone induced damage, out hypothetical bespoke temperature reponse function may have important indications for landsurface modeling, where problems in productivity of species especially in the arctic regions occure due to PFTs which are not suitable for the climate. An automatition of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

>>>>

\begin{abstract}
Here we assess the importance of characterization of key meteorological variables and air pollution profiles on the modeled susceptibility of subarctic vegetation to ozone. Changes in climate (e.g. temperature, water availability) will manifest in both direct and indirect effects influencing the species' sensitivity to air pollution. A likely increase in frequency of persistent heatwaves in the future, will inflict heat related stress while generating favorable conditions for the formation of ambient air ozone, e.g. due to wildfires.
We have monitored ambient air ozone concentrations at the Norwegian Institute of Bioeconomy Research (NIBIO) Environment Centre Svanhovd in Northern Norway in the vegetation period 2018/19. Due to data acquisition problems, ozone concentrations for two weeks in July 2018 were missing from our record. We present a reconstruction based on probability density function with respect to the Swedish and Finnish atmospheric monitoring sites and compare with available surface ozone products for the region. Over all, ozone concentrations did not differ significantly between the two years. While temperatures and global irradiance diverged substantially from the multi annual mean precipitation varied only to some extend. 
We investigate the difference in uptake of ozone using the $\mathrm{DO_3SE}$ model, for typical vegetation at the location. We find that bespoke parameterization for subarctic vegetation is essential for an accurate estimation of Phyto-toxic Ozone Dose (POD) and an assessment of damage risk. Accurate estimates of the start and length of growing season are also crucial.
We find that temperature and its derivative vapor pressure deficit are likely the most important climate variables while the variability of ozone concentration within climatologically reasonable limits plays a minor role.
The use of subarctic parameterization suggests an improvement in risk assessment as POD values are increased in 2018; corroborated by evidence of visible injury in the Svanhovd ozone garden.
\end{abstract}

@@

To measure the risk of damage inflicted to vegetation through ozone, different metrics have been defined. Two main types can be differentiated, concentration-based and exposure-based metrics, e.g., Accumulated Over a Threshold $\mathrm{x}$ ($\mathrm{AOT_x}$) and Phyto-toxic Ozone Dose over a threshold $\mathrm{y}$ ($\mathrm{POD_y}$). Due to their simplicity, concentration-based metrics are still widely used, despite comparisons between the two methods (e.g. $\mathrm{AOT_{40}}$ with $\mathrm{POD_3}$) show that exposure-based metrics are better suited to represent the actual risk to vegetation from ozone, especially in the more climatically extreme parts of Europe \citep{GCB:Mills2010,ICP:MappingManual2017,iFJuran2018}. Most of these previous studies focused on the Mediterranean region.

@@

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $7.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring is in coincidence with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} begins about one month ahead of the estimated onset of the greening season. Based on data from SeNorge.no and the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively, we find $\left<G_\mathrm{begin}\right> = (150 \pm 14)\,\unit{days of the year}$ at Svanvik (Supplement Fig~\ref{fig:greening_season_change_Svanvik}), which translates into the end of May. This finding is in line with more extensive studies \citep[][e.g.,]{IJB:Karlsen2007,AFM:Linderholm2006}.

@@

\conclusions[Discussion and conclusions]
\label{sec:conc}
We have examplarily studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. We also assessed implications on potential ozone damage risks arising at the advant of climate change. We found that conditions for ozone formation were more favourable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occured more frequently and at higher levels in 2018. This was in particular caused by the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July).

We found that the length and timing of the growing season is crucial for risk assessment based on accumulated ozone flux metrices ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for Norway spruce in 2018 to be at least one month shorter than in 2019. The found onsets ($106/122\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for downy birch and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the onset of growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods to estimators show a later onset for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

We found representative ozone climatologies (multi-annual means) for northern Fennoscandia and Svanhovd based on surface ozone monitoring. To account for missing data from our 2018 record, we proposed a routine for gap filling of ozone observations following a Reynolds's decomposition into annomalies and climatology. We also assessed the quality of available global reanalysis (ECWMF MACC and CAM) and regional model reanalysis (Copernicus) near-surface ozone concentrations products and conclude that only the Copernicus product can sufficiently reproduce observed ozone concentrations in northern Fennoscandia.

Although high and more frequent ozone peak concentrations were observed in 2018, ozone concentrations were not significantly elevated by statistical means. It therefor occurs that weather conditions and constantly high ozone concentration are more important in risk assessment using flux-based metrices like $\mathrm{POD_y}$ than episodes of peak concentrations over shorter periodes of time. Consequently, accute cellular damage caused by very high fluxes over a short periode of time is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. susceptibility to ozone depending on leaf age, in the associated damage functions might give a more differenciated picture. For herbal and grassland species (similar for crops) a strong dependency on phenological state, e.g. high susceptibility to ozone in the reproductive state, were found \citep{EP:Bassin2004}.

To estimate potential biases in assessment of ozone induced risk on Fennoscandic vegetation, we defined hypothetical temperature response functions representing different levels of assumed acclimation to cold climates. We assume northern Fennoscandic vegetation to be more tolerant towards cold temperatures and less tolerant to heat. Based on probability density functions (PDF) of observed light and temperature, we assume that an acclimation to increase the enclosed area between the respective response functions and the PDF. These hypothetical species or PFTs still need varification by experiemental data. Our results, however, allowed for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric. In particular, we identified the following main effects on the model predicted ozone uptake with respect to the default parameterization:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake regardless of individual year,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the above,
\item an earlier assumed onset of growing season $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher light threshold $\Rightarrow$ increased/reduced uptake,
\item bespoke light threshold $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years. Only accounting for acclimation, the predicted ozone uptake is typically larger than for default parameterizations. This indicates a considerable underestimation of ozone risk for subarctic species, especially for downy birch and perennial grassland. However, taking bespoke length and timing of growing season into account the opposite might be the case under certain conditions, e.g. a later start of the growing season with respect to the ozone spring peak for downy birch annd perennial grassland. Simulations of Norway spruce with bespoke (shorter than default) growing season show only little sensitivity to acclimations and inter-annual variation of meteorological conditions. We have not assessed the effect of VPD response function here, but our climatological data show that conditions at Svanhovd are shifting towards a more VPD limited regime.

@@

Based on the methodes described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and find significant reductions for downy birch ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) with respect to estimates based on $\mathrm{POD_1}$ from default parameterizations. For Norway spruce, the results are compatible with the default parameterization estimates and are contained within the range of systematic uncertainties deduced from our bespoke response function ($2.5-3\,\unit{\%}$). These results have to be treated with care, because the biomass reduction functions have been evaluated based on central European species or PFTs and might as well be subject to biases.

At the same time, intra-specied variability even at the same location is typically non neglicable \citep{EP:Bassin2004,Amb:Girgzdiene2009}. This may explain contradictory results regarding the ozone sensitivity of natural vegetation in northern Europe. While \citet{FS:Subramanian2014} reported a modeled reduction of net primary production in Norway spruce ($4.3-15.5\,\unit{\%}$) and birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, Lithuanian forest decidous trees were observed to be more susceptible to visible damage by ozone than Norway spruce \citep{Amb:Girgzdiene2009}. Open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch were exposed to elevated ozone concentrations showed biomass reductions of both downy birch and mountain birch seedlings after one-season exposure \citep{Amb:Manninen2009}. The leaves of the mountain birch in the high ozone treatments showed necrotic stipples as early as June. As mountain birch very rapidly develops during spring and early summer, compensation for damaged leaves is difficult for this species. In contrast to mountain birches, northern Scots pine of local origin only showed invisible needle-level \chem{O_3} impacts after three years of exposure. There is no field data comparing the ozone sensitivity of birches and Scots pine from northern Europe.
In our study, we do not explicitly account for this natural distribution of individual plant traits, but the chosen hypothetical, bespoke response functions might reflect this diversity within a given species or PFT. The resiliance of northern Fennoscandic plant communities to climate change and predictions in this regard strongly depends on both the existing range of acclimations as well as the actual type of acclimation.

Cumulative ozone uptake has been reported to reduce stomatal conductance significatly and independently from photosynthesis \citep[e.g.]{BGS:Lombardozzi2012}, with photosynthesis being more strongly affacted. This might be due competing effects, as an ozone induced sluggishness \citep{SR:Hoshika2015} renders stomata inapt to fully close and leads to higher stomatal conductance than expected and thus higher ozone uptake as well as transpiration. At present, the $\mathrm{DO_3SE}$ model does not account for any ozone inflicted reduction of stomatal conductance throughout the growing season and is not coupled to photosynthesis. Hence, the presented $\mathrm{POD_y}$ may be regarded as an estimated upper limit.

Beyond the risk assessment of ozone induced damage, our hypothetical bespoke temperature reponse function may have important indications for landsurface modeling, where problems in productivity of species especially in the Arctic regions occure due to PFTs which are not suitable for the climate. We pointed out with respect to default perennial grassland parameterization that that did not show any substential stomatal conductance in 2019 which we characterized as a normal year. In termas of GPP this would mean unrealistically low carbon sequestation. An automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.
